Type Refinements for the Working Class

### 2015-12-21 paper.tex
Switch to large power object, large subobject classifier

% power object
\newcommand\Pow[1]{\wp\left(#1\right)}

@@

Our position is that these views of type theory are not in conflict, but rather
describe two distinct layers in a single, harmonious system. From this
perspective, formal type theories can do little more than negotiate matters of
grammar, and therefore may serve as a linguistic framework for mathematical
expression, being responsible for the management of variable binding and
substitution, among other things. On the other hand, mathematical objects find
\emph{meaning} in their extension within the behavioral type theory, and are
subject to an extensional equality.

@@

@@

  In fact, the complexity of the above meaning explanation is an artifact of
  the pointwise style in which we have expressed it. Considered internally to
  the copresheaf topos $\Sets^\SCtx$, a refinement is merely a section of the
  object of $\Exprs_\tau$-PERs $\PERs{\Exprs_\tau}$, defined
  internally as a subobject of $\Omega^{\Exprs_\tau\times\Exprs_\tau}$:
  \[
    \Define{
      \PERs{X}
    }{
      \{ \phi : \Omega^{X\times X}
         \mid
         symmetric(\phi)
         \land transitive(\phi)
      \}
    }
  \]

  So, $\Refines{\Upsilon}{\phi}{\tau}$ obtains just when we have
  $\Member{\phi}{\PERs{\Exprs_\tau}(\Upsilon)}$. The benefit of explaining such
  judgments as objects in the topos is that we do not need to deal with
  the complexities of quantifying over renamings, since this is implicit in the
  definition of the exponential of presheaves, where
  $\Define{\Coyoneda{\Upsilon}}{\Upsilon\inj-}$ is the co-yoneda embedding:
  \begin{align*}
    \ADefine{
      B^A(\Upsilon)
    }{
      \Hom{\Sets^\SCtx}{
        \Coyoneda{\Upsilon}\times A
      }{
        B
      }
    }\\
    &\cong
    \OMode{
      \int_{\Upsilon'}
      (\Upsilon\inj\Upsilon')
      \implies
      {B(\Upsilon')}^{A(\Upsilon')}
    }
  \end{align*}

@@

The intersection of a family of refinements is a kind of uniform universal
quantification, in the sense that the terms themselves bear no trace of the
quantification, but are guaranteed in the metalogic to be related in all fibres
of $\psi$.

In practice, the family intersection is perhaps the most-used type constructor
in Nuprl developments, as it facilitates the simultaneous specification and
implementation of programs in type theory, without causing details of the
program's verification to leak into the extracted algorithm. Additionally, the
family intersection can be used to implement coinductive types.

@@

The set comprehension $\MkSet{x:\phi\mid\psi}$ contains just those elements of
$\phi$ such that their fibres of $\psi$ are inhabited; this can be seen as an
``implicit'' version of existential quantification, and can be used to
simultaneously specify and implement program without the evidence for $\psi$
intruding into the extracted algorithm.

@@

\begin{remark}
  In any topos $\mathcal{E}$, for an object $X$ we can define the object of
  relations on $X$ as the exponential $\Define{\Pow{X}}{\Omega^X}$. Then, the
  data of such a relation is contained in a global section
  $\Of{R}{\Hom{\mathcal{E}}{\mathbb{1}}{\Pow{X}}}$.
\end{remark}

@@

Stated internally as an object in the topos, the meaning of parametric typehood for
$\phi$ is perhaps a bit more clear:
\[
  \Define{
    \IsType{-}{\phi}{\tau}
  }{
    \prod_{L,M,N\in\Exprs_\tau}
    {\phi(L,N)}^{L\sim_\tau M \times \phi(M,N)}
  }
\]

>>>>

\newcommand\Pow[1]{\wp\left(#1\right)}

@@

Our position is that these views of type theory are not in conflict, but rather
describe two distinct layers in a single, harmonious system. From this
perspective, formal type theories can do little more than negotiate matters of
grammar, and therefore may serve as a linguistic framework for mathematical
expression, being responsible for the management of variable binding and
substitution. On the other hand, mathematical objects find
\emph{meaning} in their extension within the behavioral type theory, and are
subject to an extensional equality.

@@

\section{Generalized Predicative Mathematical Foundations}

Before we continue, it makes sense to remark briefly on the strength of the
mathematical foundations that we are assuming for the development of our
framework. As in Martin-L\"of's meaning-theoretic judgmental method, our
apparatus is meant to be understood in a constructive and predicative
metatheory; therefore, by $\Sets$, we do not mean the classical category of
sets, but rather the intuitionistic one, which has lacks the powerset
operation.

Therefore, in our setting, $\Sets$ is not an elementary topos, and nor is any
category of $\Sets$-valued (pre)sheaves. However, for a category of classes
$\Classes$, we can express a power class operation
$\Define{\Pow{X}}{\MkSet{Y\mid Y\subseteq X, Y\in\Sets}}$ which gives the
proper class of sub\emph{sets} of a class $X$ \cite{gambino:2002}.

A category of $\Classes$-valued presheaves then is a $\Pi W M$-pretopos (a
predicative topos with dependent function objects, wellfounded trees, and
non-wellfounded trees); such a pretopos also admits a large subobject
classifier $\Omega(c)$, defined as the class of small sieves on
$c$,\footnote{A sieve $S$ on $c$ is a subobject of the Yoneda
embedding $\Yoneda{c}$, and $S$ is \emph{small} if $S(d)$ is a set for every
stage $d$.} whence we can form the large power object
$\Define{\Pow{X}}{\Omega^X}$ for any object $X$.

In practice, the shift to a predicative topos with a large power object was a
mere technical nuisance, as all the constructions we needed to perform carried
over perfectly to this setting. For the remainder of this paper, we assume the
semantic universe $\Define{\SemU}{\Classes^\SCtx}$, the category of
copresheaves on symbol contexts. Every judgment $\mathcal{J}$ which we define
in this paper should be understood as such a presheaf, which is at each stage
$\Upsilon$ the class of derivations of $\mathcal{J}$ using symbols in
$\Upsilon$; we will also make use of \emph{internal} forms of judgment, which
are sections of $\Pow{X}$ for some object $X$.

@@

  In fact, the complexity of the above meaning explanation is an artifact of
  the pointwise style in which we have expressed it. Considered internally to
  the semantic universe $\SemU$, a refinement is merely a section of the
  object of $\Exprs_\tau$-PERs $\PERs{\Exprs_\tau}$, defined
  internally as a subobject of $\Pow{\Exprs_\tau\times\Exprs_\tau}$:
  \[
    \Define{
      \PERs{X}
    }{
      \{ \phi : \Pow{X\times X}
         \mid
         symmetric(\phi)
         \land transitive(\phi)
      \}
    }
  \]

  So, $\Refines{\Upsilon}{\phi}{\tau}$ obtains just when we have
  $\Member{\phi}{\PERs{\Exprs_\tau}(\Upsilon)}$. The benefit of explaining such
  judgments as copresheaves is that we do not need to deal with
  the complexities of quantifying over renamings, since this is implicit in the
  definition of the exponential of presheaves, where
  $\Define{\Coyoneda{\Upsilon}}{\Upsilon\inj-}$ is the co-yoneda embedding:
  \begin{align*}
    \ADefine{
      B^A(\Upsilon)
    }{
      \Hom{\Sets^\SCtx}{
        \Coyoneda{\Upsilon}\times A
      }{
        B
      }
    }\\
    &\cong
    \OMode{
      \int_{\Upsilon'}
      (\Upsilon\inj\Upsilon')
      \implies
      {B(\Upsilon')}^{A(\Upsilon')}
    }
  \end{align*}

@@

The intersection of a family of refinements is a kind of uniform universal
quantification, in the sense that the terms themselves bear no trace of the
quantification, but are guaranteed in the metalogic to be related in all fibers
of $\psi$.

In practice, the family intersection is perhaps the most ubiquitous type
constructor in Nuprl developments, as it facilitates the simultaneous
specification and implementation of programs in type theory, without causing
details of the program's verification to leak into the extracted algorithm.
Additionally, the family intersection can be used to implement coinductive
types.

@@

The set comprehension $\MkSet{x:\phi\mid\psi}$ contains just those elements of
$\phi$ such that their fibers of $\psi$ are inhabited; this can be seen as an
``implicit'' version of existential quantification, and can be used to
simultaneously specify and implement program without the evidence for $\psi$
intruding into the extracted algorithm.

@@

@@

Stated internally as an object in $\SemU$, the meaning of parametric typehood for
$\phi$ is perhaps a bit more clear:
\[
  \Define{
    \IsType{-}{\phi}{\tau}
  }{
    \prod_{L,M,N\in\Exprs_\tau}
    {\phi(L,N)}^{L\sim_\tau M \times \phi(M,N)}
  }
\]
